<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Meaningful Spatial Statistics: an introduction}
\usepackage[utf8]{inputenc}
-->
# Meaningful Spatial Statistics: The `mss` Package

This vignette introduces an R package for meaningful spatial statistics,
```{r}
library(mss)
```

The package introduces three classes for spatial data: `SField`
for representing geostatistical data, `SObjects` for
point patterns, and `SLattice` for lattice (areal)
data.

## `Window`: field domain or observation windows
`SField` and `SObjects` contain a `Window`,
representing the domain of the field, or the window of observation
for a point pattern.  

The window class defines an area: 
```{r}
showClass("Window")
```
The area can be represented either by a grid, as in
```{r}
library(sp)
demo(meuse, ask = FALSE, echo = FALSE)
w2 = Window(meuse.grid) # note the grid is passed, not the polygons
```
or by a polygon or set of polygons, as in
```{r}
w1 = Window(meuse.area)
```

## `SField`: geostatistical data
SFields represent _continuous_ functions, which over a
domain $D$ have a value at every _point_ $s \in D$.
```{r}
showClass("SField")
```
### Fields from points
Observations on fields variables are often done on irregular
point locations. Such a field variable is created from a
`SpatialPointsDataFrame` object:
```{r}
sf = SField(meuse, meuse.grid)
```
where, here, `meuse.grid` defines the domain for the
field. Alternatively, _point observations_ can be gridded, and
defined as a field by
```{r}
sf = SField(meuse.grid, meuse.grid, cellsArePoints = TRUE)
```
Note that the first argument defines the point observations, and 
the second argument the domain: the first argument is treated as
a (gridded) set of points (grid cell centres), the second as a 
gridded set of areas (cells).
### Fields from grids, grid cells representing constant areas
We can alternatively assume that each grid cell represents an area
with constant point values. Such a field is defined by:
```{r}
sf = SField(meuse.grid, meuse.grid, cellsArePoints = FALSE)
```
This field is completely known, for every point location. Note
that grid cell values are _constant throughout the cell_, and are
not conceived as cell aggregate (e.g. average, or otherwise
convoluted) values. For that case, `SpatialAggregates` are used.

### Fields from lines and polygons
Sets of points can not only be defined by grid cells, but also
through lines, or polygons. For instance, points on a contour
line could form a sample of a field:
```{r}
library(maptools)
library(rgeos)
cl = ContourLines2SLDF(contourLines(as.image.SpatialGridDataFrame(
  meuse.grid["dist"])))
proj4string(cl) = CRS(proj4string(meuse.grid))
sf.lines = SField(cl, meuse.grid)
```
or, alternatively, as a set of polygons; in this example, a single
polygon with a constant value:
```{r}
pol = addAttrToGeom(meuse.area, data.frame(value = 3), FALSE)
sf.pol = SField(pol, meuse.area)
```
## `SObjects`: point patterns or objects
The `SObjects` class can be used to represent objects in a space,
where the _window of observation_ reflects the area for which all
points or objects are available:
```{r}
showClass("SObjects")
```
We can create a point pattern e.g. for the sample locations of the `meuse`
data set:
```{r}
pts = geometry(meuse)
se.pts = SObjects(pts, meuse.area)
```
but also areas can indicate entities:
```{r}
se.area = SObjects(meuse.area, meuse.area)
```

## `SLattice`: areal data
Aggregations refer to values, measured or computed as the aggregate
over a set of (point) values, e.g. the mean, maximum or minimum value.
As such, aggregations need to be defined for _sets of points_, i.e.
for lines, polygons, or grid cells:
```{r}
showClass("SLattice")
```
_Sets of points_ can be represented by anything but points: lines, grids,
and polygons:
```{r}
sa.1 = SLattice(cl)
sa.3 = SLattice(meuse.grid)
sa.2 = SLattice(pol)
```

## Three interpretations of a spatial grid
Spatial grid data can be conceived as having values for
1. points, at grid cell centres
2. constant values throughout each grid cell
3. aggregated values over the grid cell region
The difference is illustrated by querying each at two points, one
coinciding with a grid cell centre, one not coinciding. The two
points are indicated by red circles:
```{r, fig.width=9, fig.height=6}
xy = rbind(coordinates(meuse)[1,], coordinates(meuse.grid)[10,])
pts = SpatialPoints(xy, CRS(proj4string(meuse.grid)))
plot(as(meuse.grid[1:21,], "SpatialPolygons"))
plot(pts, col = 'red', pch = 16, add = TRUE)
plot(meuse.grid[1:21,], add = TRUE)
text(coordinates(pts), c("1", "2"), pos = 2)
```

Querying a spatial field where cells are points gives non-missing
values only when the point coincides with the grid cell centre:
```{r}
f1 = SField(meuse.grid, meuse.area, cellsArePoints = TRUE)
pts = SField(pts, meuse.grid)
over(pts, f1)
```

Querying a spatial field where cells are constant valued areas gives
the grid cell values of the cells where the point is in:
```{r}
f2 = SField(meuse.grid, meuse.area, cellsArePoints = FALSE)
over(pts, f2) # matches pts to the complete grid cell
```
Querying a grid where cell values are aggregations does not answer and
generates a warning:
```{r}
a = SLattice(meuse.grid)
over(pts, a)
```

## Methods for fields: interpolation

### point-to-point interpolation
We can predict to points, laid out regularly, inside the polygon of the field domain:
```{r}
m = SField(meuse, meuse.area)
v = gstat::vgm(.6, "Sph", 900, .05) # variogram model
i1 = interpolate(log(zinc)~1, m, model = v)
spplot(i1, "var1.pred")
```

To show that these really concern points, plotting for a smaller region gives
actual point symbols:
```{r, fig.width=10, fig.height=4}
spplot(i1[1:300,], "var1.pred")
```

Alternatively, we can predict values on a pre-defined grid, taken
from the domain:
```{r}
m = SField(meuse, meuse.grid)
i2 = interpolate(log(zinc)~1, m, model = v)
print(as(i2[1:3,], "data.frame"), digits = 10) # point support
```

### point-to-area interpolation (block kriging)
From point values, we can predict the polygon area mean value:
```{r}
library(rgeos)
m3 = SLattice(meuse.area)
i3 = interpolate(log(zinc)~1, m, m3, model = v) # interpolate from point TO polygon support
i3$var1.pred # a single value, estimate of mean for the whole area
mean(log(m$zinc)) # sample mean
mean(i1$var1.pred) # mean of predictions, similar to i2$var1.pred
```
or predict values for grid cell averages (block kriging):
```{r}
m4 = SLattice(meuse.grid)
i4 = interpolate(log(zinc)~1, m, m4, model = v) # interpolate from point TO feature (cell) support
print(as(i4[1:3,], "data.frame"), digits = 10) # grid cells support
spplot(i4, "var1.pred")
```

### Support: area-to-point and area-to-area kriging
Suppose we have grid data on a 3 x 3 grid, and want to estimate on a new location, indicated by a red
circle:
```{r}
set.seed(131)
gt = GridTopology(cellcentre.offset = c(0,0), cellsize = c(1,1), cells.dim = c(3,3))
sgdf = SpatialGridDataFrame(SpatialGrid(gt), 
	data.frame(r = round(runif(9, max = 10), 1)))
plot(as(sgdf, "SpatialPolygons"), axes = TRUE)
text(coordinates(sgdf), labels = sgdf[[1]])
pt.red = SpatialPoints(cbind(0.25, 0.25))
points(pt.red, col = 'red', pch = 1)
```

The first case is when the grid cell data represent
point support values at the grid cell centre, which
is identical to case `i1` above:
```{r}
sf = SField(as(sgdf, "SpatialPointsDataFrame"), sgdf)
vm = gstat::vgm(1, "Exp", 1)
i5 = interpolate(r~1, sf, SField(pt.red, sgdf), model = vm)
as(i5[1,], "data.frame")
```
The second case, grid cells represent constant area with
point support value, meaning that every point location within
a grid cell has the value of the grid cell assigned:
```{r}
sf = SField(sgdf, sgdf, cellsArePoints = FALSE)
i6 = interpolate(r~1, sf, SField(pt.red, sgdf), model = vm)
as(i6[1,], "data.frame") # does not interpolate, but queries grid cell
```
as case `i4` above, from point values we can predict a mean value for a grid cell by
```{r}
sf = SField(sgdf, sgdf, cellsArePoints = TRUE)
gt = GridTopology(cellcentre.offset = c(.25,.25), cellsize = c(1,1), cells.dim = c(1,1))
grd.red = SpatialGrid(gt)
i7 = interpolate(r~1, sf, SLattice(grd.red), model = vm)
i7@observations@data
```

### Area-to-point kriging
```{r}
sp = as(sgdf, "SpatialPolygonsDataFrame")
gf = SLattice(sp)
kr = interpolate(r~1, gf, SField(pt.red, sp), model = vm)
kr@observations[[1]]
library(gstat)
krige0(r ~ 1, sp, pt.red, vgmArea, vgm = vm) # check
```

### Area-to-area (grid-to-grid) kriging
```{r}
sp = as(sgdf, "SpatialPolygonsDataFrame")
gr = as(grd.red, "SpatialPolygons")
grd = addAttrToGeom(gr, data.frame(r=0), FALSE)
plot(as(sgdf, "SpatialPolygons"), axes = TRUE)
text(coordinates(sgdf), labels = sgdf[[1]])
plot(grd, add = TRUE, border = 'red')
kr = interpolate(r~1, gf, SLattice(grd), model = vm)
kr@observations[[1]]
krige0(r ~ 1, sp, grd, vgmArea, vgm = vm) # check
```

## Estimating density from entitities
For `SpatialEntity` objects, we can estimate the density of objects; this
method reuses `MASS:kde2d`:
```{r}
e = SObjects(meuse, meuse.area)
d = density(e, newdata = meuse.grid)
lt = list(list("sp.polygons", meuse.area, border = 'black',
	first=FALSE), list("sp.points", meuse, col = grey(.5), pch = 16))
class(d)
spplot(d, sp.layout = lt)
```

The reason why this returns a `SLattice` is that the
values computed (densities) do not correspond to quantities that are
measurable at point locations where they are registered (grid cells).
In fact, they are aggregated values for areas even much larger than
the grid cells.

## Spatial Aggregation

### for `SField` data
```{r}
m = SField(meuse["zinc"], meuse.area)
a = aggregate(m, SLattice(meuse.area), mean) # no warning
a[[1]]
a = aggregate(m, SLattice(meuse.area), sum) # warns:
a[[1]]
```

### for `SpatialEntity` data
```{r}
m = SObjects(meuse["zinc"], meuse.area)
a = aggregate(m, SLattice(meuse.area), mean) # warns:
a[[1]]
a = aggregate(m, SLattice(meuse.area), sum) # no warning
a[[1]]
```

# Meaningful warnings

Pebesma et al. ([2014](https://agu.confex.com/agu/fm14/meetingapp.cgi#Paper/3289))
make assertions A1-A7. 

## A1. 
**A prediction is meaningful, if it provides an estimate for a potential observation
Prediction on a field variable generates no warning.**

Where prediction of a `SField` varialbe generates no warning,
```{r}
m = SField(meuse, meuse.area) # data are point support
res = interpolate(log(zinc)~1, m) # interpolate from point TO point support
```
prediction of a point pattern variable generates a warning:
```{r}
m = SObjects(meuse, meuse.area)
sa = SLattice(meuse.grid)
res = interpolate(log(zinc)~1, m, sa) # interpolate point pattern variable: warns
```
## A2. 
**Summing up values over an area is meaningful, if the observed window corresponds to the target geometry (grouping predicate) of the aggregation.**

For `SpatialEntity` data, spatial aggregation over an area larger than the 
observation window generates a warning:
```{r}
x = bbox(meuse)[1,]
y = bbox(meuse)[2,]
bb = Polygon(cbind(c(x[1],x[1],x[2],x[2],x[1]), c(y[1],y[2],y[2],y[1],y[1])))
bbx = SpatialPolygons(list(Polygons(list(bb),"bbox_meuse")),
	proj4string = CRS(proj4string(meuse)))
m = SObjects(meuse["zinc"], meuse.area)
a = aggregate(m, SLattice(bbx), sum) # warns:
```
Stasch et al. [2014](http://www.sciencedirect.com/science/article/pii/S1364815213001977)
prove A1 and A2 more formally.

Further assertions in the poster:
## A3.
**Polygon data may reflect constant values over areas (geology) or means 
(population density), this matters for (further) aggregation or downsampling.**

Package `mss` distinguishes between `SField` and
`SLattice` to represent constant values, and aggregations,
respectively. Subsampling `SField` variabls does not not warn,
```{r}
area = addAttrToGeom(meuse.area, data.frame(v=1), match.ID = FALSE)
f = SField(area, meuse.area)
pts = SField(meuse[1:5,], meuse.area)
over(pts, f) # retrieve f at the points of pts:
```
but subsampling `SLattice` variables triggers a warning, and no
useful value
```{r}
a = SLattice(area)
over(pts, a)
```
as it should really disaggregate first, from area to point.

## A4.
**The same is true for grid cells.**

`SField` and `SLattice` both generalize polygon and grid data.
```{r}
f1 = SField(meuse.grid, meuse.area, cellsArePoints = TRUE)
pts = SField(meuse[1:5,], meuse.area)
over(pts, f1) # matches pts to grid cell center points:
f2 = SField(meuse.grid, meuse.area, cellsArePoints = FALSE)
over(pts, f2) # matches pts to the complete grid cell
a = SLattice(meuse.grid)
over(pts, a)
```

## A5. 
**Point pattern data often come without observation window, over which aggregation is meaningful**

Attempting to create a `SObjects` object without an observation window specified gives
a warning message:
```{r}
pts <- SObjects(meuse)
```
Providing a window that does not contain all the points gives an error:
```{r}
tr = try(pts <- SObjects(meuse, meuse[1:10,]))
attr(tr, "condition")
```

## A6. 
**Field measurements often come without a notion for which locations interpolations based on them make sense.**

```{r}
m = SField(meuse["zinc"], meuse.area)
a = interpolate(log(zinc)~1, m, SLattice(bbx)) # warns:
```

## A7. 
**File types do not inform on meaningfulness**

`SpatialPoints`, `SpatialPixels`, `SpatialGrid`, `SpatialLines`,
`SpatialPolygons` and their `*DataFrame` counterparts do not inform
whether data concern fields (geostatistical data), entities (point
patterns), or aggregations. The classes in `mss`: `SField`,
`SObjects`, and `SLattice`, do.

## References
1. E. Pebesma, C. Stasch, B. Graeler and S. Scheider, 2014.  Meaningfully Integrating Big Earth Science Data; AGU fall meeting, poster [IN33A-3757](https://agu.confex.com/agu/fm14/meetingapp.cgi#Paper/3289) [pdf](https://agu.confex.com/data/handout/agu/fm14/Paper_3289_handout_1504_0.pdf)
2. Stasch, C., S. Scheider, E. Pebesma, W. Kuhn, 2014.  Meaningful Spatial Prediction and Aggregation.  Environmental Modelling \& Software, 51, 149-165 [open access](http://www.sciencedirect.com/science/article/pii/S1364815213001977).
